Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-06-29 analysis/paper/paper.Rmd
edited references and citation

Body ornaments are usually viewed as materials that convey social meaning or representations in a culture through the body as the scene to display [@Joyce2005]. Foreign ornaments including glass beads, agate or carnelian beads, and metal objects first appeared in Taiwan around 1,800 years ago through the regional exchange network in Southeast Asia [@WangandLiu2007]. Those new types of ornaments gradually replaced the earlier jade ornamental system in Taiwan since the transition of Late Neolithic and the Iron Age. We observed a greater quantity and variety of foreign goods in the early 17th century during the European colonization period that involved the whole island in an international trade network on a large scale. Foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts usually had an impact on local societies in many parts of the world that might cause transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997;@Junker1993;@Mitchell2000;@Silliman2005). Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed that emphasize the ability of active adoptions of foreign materials, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices [@Given2004;@Mullins2011;@Silliman2001;@Rubertone2000;@Torrence2000;@TorrenceandClarke2000;@Voss2005).

Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local communities remains unanswered. Indirect effects of colonialism is addressed recently to discuss the impact on the local indigenous societies in the periphery of colonial control [@Trabert2017]. Taiwan was colonized by the Spanish and the Dutch during the early 17th century, where northeastern Taiwan, Yilan, is an especially unique example for exploring indirect effects due to its physically isolated location that made it not easily accessible to the Europeans and the colonial control was less compared to other parts of this region [@Andrade2007; @Kang2012). Prior to the contact with the Europeans, there were small-scale regional exchange networks between China and Taiwan since in the Neolithic period. Also, northern Taiwan had been involved in long-distance networks of East Asia since 14th century, and encountered a large wave of Han Chinese migration in the 19th century. Northeastern Taiwan can provide a good example to explore the indirect impacts on indigenous societies caused by different foreign groups in different periods by taking a long-term perspective research.

We ask whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan. Our hypothesis is that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network. We hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan. We hypothesize that there are clustered distribution of ornaments at Kiwulan relating to these international contacts.

@@

Although the northeastern Taiwan forms a circumscribed environment due to the surrounding mountains and sea, the sea provides a way through which indigenous people could interact with other groups from different regions, including other native indigenous groups in northern and southern Taiwan, foreign traders such as the Han people of China, and Europeans. This local exchange network of northeastern Taiwan had global connections after the 16th century because of its location on shipping routes between China and Japan (Chen 2005; Wang and Liu 2007). The groups involved in the exchange network consisted by the Chinese traders, indigenous groups called Basay in northern Taiwan, and indigenous groups, Kavalan, in northeastern Taiwan. Historical records indicate that the Kavalan people were one of the Basay’s trading partners, and they exchanged rice, deer skin, and gold for metal tools or ornaments, beads, and Chinese ceramics (Chen 2005; Hsieh 2009a; Li and Wu 2006).

Historical records from Spanish and Dutch provide rich information about the historical and cultural background both for Europeans and indigenous people in the early 17th century. Spanish occupied and built forts in Northern Taiwan since 1624, from which Spanish missionaries often went to indigenous settlement and left behind numerous records from those places. The report of the Dominican Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) as a materials to exchange with necessities they need with other indigenous people. This trend then gradually spread to other areas, and even the Spanish soldiers used agate as bargaining chips for gambling, since agate equaled money (Li 2006: 132-149).  European contact was focused on trading with indigenous communities in northeastern Taiwan. Trading activity of many types of goode among these indigenous communities increased with the arrival of European colonisers. Thus, we predict that this trading activity resulted in an increase in the amount and diversity of ornaments coming into indigenous communities in northeastern Taiwan.

Father Jacinto Esquivel recorded how people in Tamsui entered marriages. A man who wanted to take a wife had to give her parents cuentas. If a married man was found having cheated on his wife, he only had to pay the witness with agate or golden beads and the matter was settled [@Li2006:151]. Other records mention that the female shamans (majuorbol) in the tribe used agate beads for healing ceremonies in rituals [@Pao 2008: 122, 143, 151]. Regarding funerals, some indigenous people buried the dead under their houses or in the vicinity. They put a mat into the grave to protect the body from moisture and then put cooked rice on both sides of the head as provision. The deceased were buried in small graves with both knees bent, and on the grave they placed quivers with arrows, pottery, agate beads and other things that they would need. The greater the influence of the deceased, the more agate, pottery and cloths were placed on the grave [@Li 2006: 153].

Documents from the Qing dynasty usually describe the decorative purposes of the ornaments. In the Kavalan Subprefecture Gazetteer [@Chen2006[1852]: 308], the Kavalan Zhi Lue [@Ko1993[1837]: 11, 126) and the Dong Cha Ji Lue [@Yao1996[1829]: 77], it is mentioned that aborigines in Yilan used metal threads to weave so called golden carp-shaped ornaments. Their number was small and their value high, hence only the rich possessed them. Ordinary people wore agate beads or glass beads on their head or neck during various festivals. In 1895, at the beginning of Japanese rule over Taiwan, Ino Kanori [@Kanori1996: 227-232] made field surveys among various ethnic groups in Taiwan. He describes the Plains Aborigines from Yilan as not using carp-shaped ornaments any more, but mentions that older people still use beads ornaments.

Recent ethnographic research of the Kavalan tribe describes agate beads as used in divination ceremonies, the subli, by female shamans [@Liu2008: 133-134]. Interview records describe agate beads as valuable objects passed down from mother to daughter. When they were not used, they were often put on roof beams so that they were not seen by other people. But the origin of these beads is still not known. Later, because they were bought by Japanese and people in Taipei, agate beads became rarer and rarer [@Hsu1992: 22] and carp-shaped ornaments had completely disappeared. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination. It was usually the female shamans that were responsible for divination rites, a technique passed down by ancestors. Most aborigines today don’t know the origin of agate beads any more, but their worth is still highlighted by oral history and through their scarcity.

@@

The earliest record of direct European contact with indigenous people in Yilan can be traced back to 1632, when the local settlements were attacked by the Spanish according to the official documents (add citation; Borao Mateo 2009). Later in 1647, the Dutch attacked the indigenous villages in Yilan and forced them to accept colonial rules and economic demands by paying annual tribute (add citation; Borao Mateo 2009). According to Dutch census reports in 1650, Kiwulan was the biggest indigenous community in the Yilan Plain, with a population of 840 adults (Chen 2007; Kang 2012; Li 2014). The diagnostic artifacts during European contact found at Kiwulan including An-ping jars and glass beads that were largely introduced to Taiwan during the early 17th century.

The European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The evidence of Chinese migration can be identified from Chinese official records [b]and large amount of Chinese blue-and-white porcelain found at Kiwulan.

@@

The ornaments studied in this paper came from 40 adjacent 4m x 4m units out of 262 units in the middle part of excavation. The sampling principle is based on the extent of intact contexts that the least disturbed by modern constructions are the preference. The Upper Layer Culture can be divided into six sub layers spanning from 14 century to 19 century according to current radiocarbon dates, excavation depth, types of ceramics, and sediment texture and color (Hsieh 2008; Wang 2011). Layers are assigned numbers from 1 to 6 for which L1 corresponds to the upper context and L6, also the bottom layer, refers to lower context. Based on Hsieh’s (2008) chronological study of Chinese porcelains, L5 and L6 represent pre-European contact period, L4 was the time of European contact, and L2 and L1 was the period of Chinese contact.

@@

In general, complete ornaments with clear context were mostly found in burials that helps to understand how people wear them and any relationships between shapes of ornaments and gender or age. In addition to burials, ornaments were also found in middens and living space that can be identified by post holes. The frequency of ornaments from burials is more than those from the living space due to well preserved condition. There is a wide variety of raw materials for ornaments of which ornaments glass, agate and metal ornaments that account for 99% of all ornaments were thought to be introduced by trading with the Europeans and Chinese traders, while ornaments made of shells, woods, and bones were believed locally made that is about only about 1% of all ornaments (Table 1). The reason for the large quantity of glass beads is that they usually have been found in clusters because of the original form such as necklaces. The shape of ornaments include beads, bracelets, rings, bells, pendants, and knitted objects that shows shape are closely related to raw material due to its characteristic. For example, beads are only made of glass, agate, and shell, and bone is the main material for knitted objects that used to connect beads. Also, metal ornaments covers more shapes such bells, pendants, and rings (Chen 2007).

@@

## Spatial distribution
Point pattern analysis is a way to assess whether the distribution of ornaments presents some spatial pattern that can indicate the changes during culture contact period. The location where ornaments from were square-based recorded, thus we took the intensity approach that focuses on the average density of points across space to explore their distribution. We randomly assigned points, here the location of ornaments, for each square according to the data, subset three groups for three time periods, and compute the kernal density for each time period for comparison. Kernal density estimations (KDE) can estimate the probability of the density of events across space by creating a continuous, smooth density surface across space. Here we use KDE to visualize the event of the presence of ornaments that were represents by maps on which we are able to see core areas and surrounding neighborhoods [@Bonnier.etal]. Density value is assigned to each cell, of which the unit is meter. The results \@ref(fig:plot-all-densities) show that there is one major core area during the pre-European contact period, multiple core areas during European contact period, and a single core during the Chinese contact period. Those maps present three consistent sub-region with shifting core area over time. The distribution might indicate different social groups who own more ornaments, and the multiple groups during European contact period might reflect a more corporate mode of social organization. However, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create obvious core area. Next step is to test whether the distribution of ornaments is randomly distributed to understand the underlying social indication related to ornaments.

@@

The introduction of ornaments to Northeastern Taiwan during different periods could inform how local indigenous people perceive and distribute the trading goods in different contexts of culture contacts by examining their frequency and spatial distribution across space. Taiwan island had been involved in regional exchange networks of Southeast Asia since Neolithic period, a global trade network during the European contact period in the early 17th century, and later intense but smaller scale network with Han Chinese communities since 19th century [@Chen200?]. Kiwulan demonstrates continuous occupation from the late Iron Age to one hundred years ago over 600 years that covers these three different culture contacts that enables a comparative research to explore the indirect influence of culture contact especially in the global colonial context. These three analytic chronology at Kiwulan represent different scale of exchange network and intensity of culture contact. For the scale of trade network, Pre-European contact can be defined as a middle scale represented by the prehistoric regional trade network in Southeast Asia. International trade network during the European contact presents a global scale, while local trade network during the Chinese contact presents a small scale. For the intensity of culture contact, both pre-European contact and European contact period could be defined as indirect contact, but the colonial power underlying the contact with the Europeans should be also considered. The culture contact during the Chinese period is more direct due to the migration of Han Chinese to Kiwulan.

>>>>

Body ornaments are usually viewed as materials that convey social meaning or representations in a culture through the body as the scene to display [@Joyce2005]. Foreign ornaments including glass beads, agate or carnelian beads, and metal objects first appeared in Taiwan around 1,800 years ago through the regional exchange network in Southeast Asia [@Wang2007]. Those new types of ornaments gradually replaced the earlier jade ornamental system in Taiwan since the transition of Late Neolithic and the Iron Age. We observed a greater quantity and variety of foreign goods in the early 17th century during the European colonization period that involved the whole island in an international trade network on a large scale. Foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts usually had an impact on local societies in many parts of the world that might cause transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed that emphasize the ability of active adoptions of foreign materials, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices [@Given2004; @Mullins2011; @Silliman2001; @Rubertone2000; @Torrence2000; @Torrence2000a; @Voss2005].

Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local communities remains unanswered. Indirect effects of colonialism is addressed recently to discuss the impact on the local indigenous societies in the periphery of colonial control [@Trabert2017]. Taiwan was colonized by the Spanish and the Dutch during the early 17th century, where northeastern Taiwan, Yilan, is an especially unique example for exploring indirect effects due to its physically isolated location that made it not easily accessible to the Europeans and the colonial control was less compared to other parts of this region [@Andrade2007; @Kang2012]. Prior to the contact with the Europeans, there were small-scale regional exchange networks between China and Taiwan since in the Neolithic period. Also, northern Taiwan had been involved in long-distance networks of East Asia since 14th century, and encountered a large wave of Han Chinese migration in the 19th century. Northeastern Taiwan can provide a good example to explore the indirect impacts on indigenous societies caused by different foreign groups in different periods by taking a long-term perspective research.

We ask whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan[@Chen2007]. Our hypothesis is that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network. We hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan. We hypothesize that there are clustered distribution of ornaments at Kiwulan relating to these international contacts.

@@

Although the northeastern Taiwan forms a circumscribed environment due to the surrounding mountains and sea, the sea provides a way through which indigenous people could interact with other groups from different regions, including other native indigenous groups in northern and southern Taiwan, foreign traders such as the Han people of China, and Europeans. This local exchange network of northeastern Taiwan had global connections after the 16th century because of its location on shipping routes between China and Japan [@Chen2005; @Wang2007]. The groups involved in the exchange network consisted by the Chinese traders, indigenous groups called Basay in northern Taiwan, and indigenous groups, Kavalan, in northeastern Taiwan. Historical records indicate that the Kavalan people were one of the Basay’s trading partners, and they exchanged rice, deer skin, and gold for metal tools or ornaments, beads, and Chinese ceramics [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical records from Spanish and Dutch provide rich information about the historical and cultural background both for Europeans and indigenous people in the early 17th century. Spanish occupied and built forts in Northern Taiwan since 1624, from which Spanish missionaries often went to indigenous settlement and left behind numerous records from those places. The report of the Dominican Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) as a materials to exchange with necessities they need with other indigenous people. This trend then gradually spread to other areas, and even the Spanish soldiers used agate as bargaining chips for gambling, since agate equaled money [@LiandWu2006, pp. 132-149].  European contact was focused on trading with indigenous communities in northeastern Taiwan. Trading activity of many types of goode among these indigenous communities increased with the arrival of European colonisers. Thus, we predict that this trading activity resulted in an increase in the amount and diversity of ornaments coming into indigenous communities in northeastern Taiwan.

Father Jacinto Esquivel recorded how people in Tamsui entered marriages. A man who wanted to take a wife had to give her parents cuentas. If a married man was found having cheated on his wife, he only had to pay the witness with agate or golden beads and the matter was settled [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe used agate beads for healing ceremonies in rituals [@Pao2008, pp. 122-151]. Regarding funerals, some indigenous people buried the dead under their houses or in the vicinity. They put a mat into the grave to protect the body from moisture and then put cooked rice on both sides of the head as provision. The deceased were buried in small graves with both knees bent, and on the grave they placed quivers with arrows, pottery, agate beads and other things that they would need. The greater the influence of the deceased, the more agate, pottery and cloths were placed on the grave [@LiandWu2006, pp. 153].

Documents from the Qing dynasty usually describe the decorative purposes of the ornaments. In the Kavalan Subprefecture Gazetteer [@Chen2006[1852]: 308], the Kavalan Zhi Lue [@Ko1993[1837]: 11, 126) and the Dong Cha Ji Lue [@Yao1996[1829]: 77], it is mentioned that aborigines in Yilan used metal threads to weave so called golden carp-shaped ornaments. Their number was small and their value high, hence only the rich possessed them. Ordinary people wore agate beads or glass beads on their head or neck during various festivals. In 1895, at the beginning of Japanese rule over Taiwan, Ino Kanori [@Kanori1996, pp. 227-232] made field surveys among various ethnic groups in Taiwan. He describes the Plains Aborigines from Yilan as not using carp-shaped ornaments any more, but mentions that older people still use beads ornaments.

Recent ethnographic research of the Kavalan tribe describes agate beads as used in divination ceremonies, the subli, by female shamans [@Liu2008, pp. 133-134]. Interview records describe agate beads as valuable objects passed down from mother to daughter. When they were not used, they were often put on roof beams so that they were not seen by other people. But the origin of these beads is still not known. Later, because they were bought by Japanese and people in Taipei, agate beads became rarer and rarer [@Hsu1992, pp. 22] and carp-shaped ornaments had completely disappeared. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination. It was usually the female shamans that were responsible for divination rites, a technique passed down by ancestors. Most aborigines today don’t know the origin of agate beads any more, but their worth is still highlighted by oral history and through their scarcity.

@@

The earliest record of direct European contact with indigenous people in Yilan can be traced back to 1632, when the local settlements were attacked by the Spanish according to the official documents [add citation; Borao Mateo 2009]. Later in 1647, the Dutch attacked the indigenous villages in Yilan and forced them to accept colonial rules and economic demands by paying annual tribute [add citation; Borao Mateo 2009]. According to Dutch census reports in 1650, Kiwulan was the biggest indigenous community in the Yilan Plain, with a population of 840 adults [@Chen2007; @Kang2012; @Li2014]. The diagnostic artifacts during European contact found at Kiwulan including An-ping jars and glass beads that were largely introduced to Taiwan during the early 17th century.

The European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The evidence of Chinese migration can be identified from Chinese official records and large amount of Chinese blue-and-white porcelain found at Kiwulan.

@@

The ornaments studied in this paper came from 40 adjacent 4m x 4m units out of 262 units in the middle part of excavation. The sampling principle is based on the extent of intact contexts that the least disturbed by modern constructions are the preference. The Upper Layer Culture can be divided into six sub layers spanning from 14 century to 19 century according to current radiocarbon dates, excavation depth, types of ceramics, and sediment texture and color [@Hsieh2009; @Wang2011]. Layers are assigned numbers from 1 to 6 for which L1 corresponds to the upper context and L6, also the bottom layer, refers to lower context. Based on @Hsieh2009 chronological study of Chinese porcelains, L5 and L6 represent pre-European contact period, L4 was the time of European contact, and L2 and L1 was the period of Chinese contact.

@@

In general, complete ornaments with clear context were mostly found in burials that helps to understand how people wear them and any relationships between shapes of ornaments and gender or age. In addition to burials, ornaments were also found in middens and living space that can be identified by post holes. The frequency of ornaments from burials is more than those from the living space due to well preserved condition. There is a wide variety of raw materials for ornaments of which ornaments glass, agate and metal ornaments that account for 99% of all ornaments were thought to be introduced by trading with the Europeans and Chinese traders, while ornaments made of shells, woods, and bones were believed locally made that is about only about 1% of all ornaments (Table 1). The reason for the large quantity of glass beads is that they usually have been found in clusters because of the original form such as necklaces. The shape of ornaments include beads, bracelets, rings, bells, pendants, and knitted objects that shows shape are closely related to raw material due to its characteristic. For example, beads are only made of glass, agate, and shell, and bone is the main material for knitted objects that used to connect beads. Also, metal ornaments covers more shapes such bells, pendants, and rings [@Chen2007].

@@

## Spatial distribution
Point pattern analysis is a way to assess whether the distribution of ornaments presents some spatial pattern that can indicate the changes during culture contact period [@Ducke2015]. The location where ornaments from were square-based recorded, thus we took the intensity approach that focuses on the average density of points across space to explore their distribution. We randomly assigned points, here the location of ornaments, for each square according to the data, subset three groups for three time periods, and compute the kernal density for each time period for comparison. Kernal density estimations (KDE) can estimate the probability of the density of events across space by creating a continuous, smooth density surface across space. Here we use KDE to visualize the event of the presence of ornaments that were represents by maps on which we are able to see core areas and surrounding neighborhoods [@Bonnier2019, @Cortegoso2016]. Density value is assigned to each cell, of which the unit is meter. The results \@ref(fig:plot-all-densities) show that there is one major core area during the pre-European contact period, multiple core areas during European contact period, and a single core during the Chinese contact period. Those maps present three consistent sub-region with shifting core area over time. The distribution might indicate different social groups who own more ornaments, and the multiple groups during European contact period might reflect a more corporate mode of social organization. However, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create obvious core area. Next step is to test whether the distribution of ornaments is randomly distributed to understand the underlying social indication related to ornaments.

@@

The introduction of ornaments to Northeastern Taiwan during different periods could inform how local indigenous people perceive and distribute the trading goods in different contexts of culture contacts by examining their frequency and spatial distribution across space. Taiwan island had been involved in regional exchange networks of Southeast Asia since Neolithic period, a global trade network during the European contact period in the early 17th century, and later intense but smaller scale network with Han Chinese communities since 19th century [@Chen2005]. Kiwulan demonstrates continuous occupation from the late Iron Age to one hundred years ago over 600 years that covers these three different culture contacts that enables a comparative research to explore the indirect influence of culture contact especially in the global colonial context. These three analytic chronology at Kiwulan represent different scale of exchange network and intensity of culture contact. For the scale of trade network, Pre-European contact can be defined as a middle scale represented by the prehistoric regional trade network in Southeast Asia. International trade network during the European contact presents a global scale, while local trade network during the Chinese contact presents a small scale. For the intensity of culture contact, both pre-European contact and European contact period could be defined as indirect contact, but the colonial power underlying the contact with the Europeans should be also considered. The culture contact during the Chinese period is more direct due to the migration of Han Chinese to Kiwulan.
